---
output: github_document
always_allow_html: true
---

```{r, include=F, echo=F}
devtools::load_all()
library(tidyverse)
```

# Gaussian Process Regression (GPR) in R

`zoomerGP` is an R package that allows you to compose and fit Gaussian Process
Regression models to data. The core inference code is implemented in Rust for
speed, and approximations are available to allow users to fit models to large
datasets (N > 10,000). 

## Supported Kernel Functions:

The flagship feature of this package is a domain-specific language that helps
you easily compose complicated kernels out of basic building blocks using a
formula syntax.

As an example, you can specify an rbf (gaussian) kernel that acts on a single
variable, x, with the following syntax:

```{r, eval=F}
gaussian_process(y ~ rbf(x), data = data)
```

If you want to regress Y on an ARD-RBF kernel that acts on all columns of the
input, you can simply use `rbf()`:

```{r, eval=F}
gaussian_process(y ~ rbf(), data = data)
```

This is because any kernel function called with no arguments will select all
columns by default. Additionally, please note that all kernels are Automatic
Relevance Determination (ARD) kernels, so a separate bandwidth is fit for each
input dimension. This generally delivers superior performance in higher
dimensions as it allows the regression to 'zero out' or not respond to input
variables that do not matter.

You can also combine kernels by adding and multiplying them with `+` and `*`.
To give an example, we can fit a linear kernel times an RBF kernel with the
following code:

```{r, eval=F}
gaussian_process(y ~ linear(x) * rbf(x), data = data)

## adding in another two predictors, a and b that enter seperately
gaussian_process(y ~ linear(x) * rbf(x) + rbf(a,b), data = data)
```

`+` and  `*` can be arbitrarily chained to combine as many kernels as you would
like. If you would like to fit a very complex function, consider using the
spectral mixture (`spectral[0-9]()`) kernels, which can generalize any
stationary covariance function. These are great for fitting complex, or
periodic functions (esp. time series).

You can also see a list of currently implemented kernels below. Formulas for
the kernels and more features are forthcoming:


| Kernel                     | DSL Name            |
| :--------------------------|:-------------------:|
| Squared Exponential        | rbf()               |
| Spectral Mixture           | spectral[0-9]()     |
| Linear                     | linear()            |
| Indicator                  | indicator()         |

## Fitting to Large Datasets with Sparse Approximations.

Gaussian Process Regression are computationally intensive, and generally takes
$O(n^3)$ computations to fit. To allow the method to scale to large datasets,
this package implements the projected process approximation as described in
Rassmusen and Williams's (2005) Gaussian Processes for Machine Learning. To
turn this approximation on, use the `sparse=T` argument when fitting the
gaussian processes please be aware that you may also need to change (reduce)
the default value of the number of inducing points (`n_points`) for numerical
stability.

## Optimization Methods

Currently, we the package supports 3 engines for hyperparameter optimization.
`bfgs` is selected by default, but these can be changed using the
training_method argument to `gaussian_process` function

* `bfgs` - as implimented in the optim package in R. Will sometimes fail to
  coverge, leading to an error.
* `rgenoud` - as implimented in the `rgenoud package.` This is likely the most
  robust optimizer as it avoids getting stuck in local minima, but also the
  most computationally intensive.
* `coin` - a learning-rate free algorithm that is implimented in Rust. Seems to
  work well across a variety of problems, but can get stuck in local minima.


## Feature Roadmap

The GPR package is a work in progress. This section summarizes the progress
towards important feature goals we hope to include in the package:

- [x] Type-2 ML estimation of hyperparameters (empirical Bayes) for Gaussian likelihood.
- [x] Sparse approximations to Gaussian likelihood
- [x] Implement [spectral mixture kernels](https://arxiv.org/pdf/1302.4245) to allow for more expressive modelling.
- [x] [COIN-optimizer](https://arxiv.org/abs/1705.07795) for fast fitting
- [] Implement derivatives of GP's and average marginal effects (A.M.E.'s)
- [] Support for non-Gaussian likelihoods via variational Bayes.

### Installing Rust

If your operating system or version of R is not installed, you must have the
[Rust compiler](https://www.rust-lang.org/tools/install) installed to compile
this package from sources. After the package is compiled, Rust is no longer
required, and can be safely uninstalled.

#### Installing Rust on Linux or Mac:

To install Rust on Linux or Mac, you can simply run the following snippet in
your terminal.

``` sh
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh
```

#### Installing Rust on Windows:

To install Rust on windows, you can use the Rust installation wizard,
`rustup-init.exe`, found [at this
site](https://forge.rust-lang.org/infra/other-installation-methods.html).
Depending on your version of Windows, you may see an error that looks something like this:

```
error: toolchain 'stable-x86_64-pc-windows-gnu' is not installed
```

In this case, you should run `rustup install stable-x86_64-pc_windows-gnu` to
install the missing toolchain. If you're missing another toolchain, simply type
this in the place of `stable-x86_64-pc_windows-gnu` in the command above.

### Installing Package from Github:

Once you have rust installed Rust, you should be able to install the package
with either the install.packages function as above, or using the
`install_github` function from the `devtools` package or with the `pkg_install`
function from the `pak` package.

``` r
## Install with devtools
# install.packages("devtools")
devtools::install_github("beniaminogreen/GPR")

## Install with pak
# install.packages("pak")
pak::pkg_install("beniaminogreen/GPR")
```

### Loading The Package

Once the package is installed, you can load it into memory as usual by typing:

```{r, warning = FALSE, message = FALSE, eval = F}
library(zoomerGP)
```
